Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-11 aistats.tex
Correct problem posing
The order in which I brought up the concepts was not so logical.

\usepackage[capitalise]{cleveref}
\newcommand{\RLP}[1]{\textcolor{red}{RLP:#1}}
\newcommand{\fdk}[1]{\textcolor{Periwinkle}{fdk:#1}}
\newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}

@@

@@

Exponential Families are to statistics what linear models are to regression.
Many classical model are exponential families, and their convexity and modeling flexibility make them one of the most widely used statistical models.
Here we review some of their properties.
We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
The exponential family for data $X\in \cX$ with natural (or primal) parameter $\nat$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
\begin{equation}
	 p(X|\nat) = \exp( \nat^\top T(X) - \logpart(\nat)) \; ,
\end{equation}
Here $\logpart$ is the log-partition function -- i.e. the normalization term
\begin{align}
    \logpart(\nat) = \log \int e^{\nat^\top T(x)} \nu(dx) \; .
\end{align}
When explicit, we will write the random variable $T = T(X)$ for lightness.
For convenience, we focus on steep, regular exponential families
where $\logpart$ is strictly convex function of Legendre type,
and the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex
(See \citet{barndoffnielsen2014information}).
This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

@@

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat)
    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
    &\propto \exp(n_0 \m_0^\top \nat - n_0 \logpart(\nat))
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
Intuitively, $n_0$ is a number of fictive data points observed from a distribution with parameter $\nat_0$.

@@

\section{Open Problem}

Let $\nat^* := \argmin \expect[X\sim\cD]{-\log p(X|\nat)}$, where $\cD$ is any distribution such that $\E[T] = \meanp^* \in \cM$.
Then the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{multline}
    \expect[X\sim \cD]{-\log p(X|\nat) + \log p(X|\nat^*) } \\
	= \KL( p(.|\nat^*) ; p(.|\nat)) \; .
\end{multline}
Note that $\cD$ does not have to be part of the exponential family.
For the exponential family, the KL is equal to the Bregman divergence induced by the log-partition function, or alternatively by the entropy, being careful with the order of the arguments
\begin{align}
\boxed{
	\KL( p(.|\nat^*) ; p(.|\nat))
    = \bregman (\nat ; \nat^*)
    = \bregmanconj ( \meanp^* ; \meanp)
}
\end{align}

Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \TODO{cite}
Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \TODO{cite}
A contrario the KL is not smooth, and it can be infinite in misspecified settings.
But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.

Our question is: how does the KL behave when $\nat$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities ?
\begin{align}
	\label{eq:bregmanMLE}
	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
	\label{eq:bregmanMAP}
	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )}
\end{align}
where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
More explicitly, can we write an upper bound that does not involve an expectation over the dataset -- e.g. where the stochasticity is removed.

\paragraph{Remark.}
This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
However there is a connection between the random variable $T(X)$ and the metric $\logpart$.
Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable.
For instance, if we plug in $\conj(\m)= -\log(\m)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.

\paragraph{Remark 2.}
The expectation of the MLE may be infinite, for instance with $\cN(0,\sigma^2)$ and $n\leq 2$. Instead of taking the expectation,  we might want to bound this quantity in high probability, without resorting to Markov inequality, but that is a difficult endeavor.

@@

\subsection{Locally Quadratic Case}
All Bregman divergences are locally quadratic.
Under some assumptions, we can quantify  when this quadratic behavior kicks in.
For instance {\bf when $\conj$ is self-concordant} \citep[Ch.4.1]{nesterov2003introductory}.
This is true when $T$ is 1-dimensional and $\logpart$ is self-concordant
-- e.g. exponential distribution,
Gaussian\fdk{with known variance?} and Laplace distributions with known mean,
Pareto distribution with known minimum value,
Weibull distribution with known shape $k$
-- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
\begin{align}
	 \bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF^*}^2
\end{align}
holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$.\footnote{The value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\nat^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.

@@

\paragraph{Illustrations.}
We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}. \RLP{Tune figures.}

@@

\newpage
\section{Notation convention}
We use
\begin{itemize}
	\item $\theta, \nat$ for natural parameters
	\item $\m$ for mean parameters
	\item $X_1,\ldots,X_n$ for data
	\item $\cX$ for the data space
	\item $T$ for the sufficient statistics
	\item $\nu$ for the base measure
	\item $A, A^*$ for the log-partition and its dual
	\item $\cB_A$ for the Bregman divergence induced by $A$
	\item $n$ for the number of samples
	\item $(\theta^*,m^*)$ for the optimum
	\item $\gamma$ for the step-size
	\item $\mu,\sigma^2$ for the parameters of a Gaussian
	\item $\hat\theta_n$ for estimates
	\item $\tilde\theta_n$ for $\expect{\hat\theta_n}$, which is biased
	\item $(L, \mu)$ for strong-convexity and smoothness
\end{itemize}

>>>>

\usepackage[capitalise]{cleveref}
\newcommand{\rlp}[1]{\textcolor{red}{RLP:#1}}
\newcommand{\fdk}[1]{\textcolor{Periwinkle}{fdk:#1}}
\newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}
\newcommand{\tocite}{\textcolor{purple}{(add citation)}}

@@

Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \tocite
Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \tocite 
A contrario the KL is not smooth, and it can be infinite in misspecified settings.
But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.

@@

Exponential Families are to statistics what linear models are to regression.
Many classical model are exponential families, and their convexity and modeling flexibility make them one of the most widely used statistical models.
Here we review some of their properties.
We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
The exponential family for data $X\in \cX$ with natural (or primal) parameter $\nat$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
\begin{equation}
	 p(X|\nat) = \exp( \langle \nat, T(X) \rangle - \logpart(\nat)) \; ,
\end{equation}
where $\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.
Here $\logpart$ is the log-partition function -- i.e. the normalization term
\begin{align}
    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \; .
\end{align}
When explicit, we will write the random variable $T = T(X)$ for lightness.
For convenience, we focus on steep, regular exponential families
where $\logpart$ is strictly convex function of Legendre type,
and the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex
(See \citet{barndoffnielsen2014information}).
This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

@@

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat)
    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat))
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
Intuitively, $n_0$ is a number of fictive data points observed from a distribution with parameter $\nat_0$.

@@

\section{Open Problem}
Given a distribution $\cD$, our objective is the expected log-likelihood 
\begin{align}
	f(\nat) = \expect[X \sim \cD]{-\log p(X \cond \nat)} 
	 = \logpart(\nat) - \langle \E[T] , \nat \rangle
\end{align}
Up to a constant, it is also the KL  $f(\nat) = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$.
Note that $\cD$ does not have to be part of the exponential family.
Let $\nat^* := \argmin f(\nat)$.
Assuming $\nat^*$ exists within $\Theta$ means $\cD$ is such that $\E[T] = \meanp^* \in \cM$.
Then the suboptimality on the population log-likelihood can be written with Bregman divergences
\footnote{
Since for exponential families, $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, this proves on the way the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
}
\begin{align}
	 f(\nat) - f(\nat^*)
	 %&= \logpart(\nat) - \logpart(\nat^*) - \langle \E[T] , \nat - \nat^* \rangle \\
	 = \bregman(\nat ; \nat^*) 
	 = \bregmanconj(\m^* ; \m) \; .
\end{align}
Our question is: how does this suboptimality behave when $\nat$ is the MLE or the MAP ? Can we get upper-bounds in expectations ? e.g. on these quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
	\label{eq:bregmanMAP}
	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )}
\end{align}
where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
More explicitly, can we write an upper bound that does not involve this expectation over the dataset.

In general, for arbitrary distributions $\cD$, it is a hard problem, especially since the KL is known for being brittle to misspecification \tocite, so let us focus on the well-specified case first $\cD = p(\cdot \cond \nat^*)$. Surely this problem has been solved, hasn't it ?
In some special cases yes, but it seems the general answer is no. 
We are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n$.

{\bf Remark.}
This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
However there is a connection between the random variable $T(X)$ and the metric $\logpart$.
Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable.
For instance, if we plug in $\conj(\m)= -\log(\m)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.

{\bf Remark 2.}
The expectation of the MLE may be infinite, for instance with $\cN(0,\sigma^2)$ and $n\leq 2$. Instead of taking the expectation,  we might want to bound this quantity in high probability, without resorting to Markov inequality, but that is a difficult endeavor.

@@

\subsection{Locally Quadratic Case}
All Bregman divergences are locally quadratic.
Under some assumptions, we can quantify  when this quadratic behavior kicks in.
For instance {\bf when $\conj$ is self-concordant} \citep[Ch.4.1]{nesterov2003introductory}.
This is true when $T$ is 1-dimensional and $\logpart$ is self-concordant
-- e.g. exponential distribution,
Gaussian with known mean,
Laplace with known mean,
Pareto with known minimum value,
Weibull with known shape $k$
-- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
\begin{align}
	 \bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF^*}^2
\end{align}
holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$.\footnote{The value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\nat^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.

@@

\paragraph{Illustrations.}
We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}. \rlp{Tune figures.}

@@

\newpage
\section{Notation convention}
We use
\begin{itemize}
	\item $\theta, \nat$ for natural parameters
	\item $\m$ for mean parameters
	\item $X_1,\ldots,X_n$ for data
	\item $\cX$ for the data space
	\item $T$ for the sufficient statistics
	\item $\nu$ for the base measure
	\item $A, A^*, \logpart, \conj$ for the log-partition and its dual
	\item $\cB_A$ for the Bregman divergence induced by $A$
	\item $n$ for the number of samples
	\item $(\theta^*,m^*)$ for the optimum
	\item $\gamma, \lr$ for the step-size
	\item $\mu,\sigma^2$ for the parameters of a Gaussian
	\item $\hat\theta_n$ for estimates
	\item $\tilde\theta_n$ for $\expect{\hat\theta_n}$, which is biased
	\item $(L, \mu)$ for strong-convexity and smoothness
\end{itemize}
